---
title: "Untitled"
author: "Gracie Lovell"
date: "2026-02-12"
output: pdf_document
---

```{r}
library(readr)
library(dplyr)
library(caret)
library(car)
nfl_cleaned <- read_csv("nfl_cleaned.csv")
head(nfl_cleaned)
```

```{r}
off_vars <- nfl_cleaned %>%
  select(Points_Scored, Off_First_Downs, Off_Total_Yards, Off_Passing_Yards, Off_Rushing_Yards, Turnovers_Lost)
head(off_vars)

def_vars <- nfl_cleaned %>%
  select(Points_Allowed, Def_First_Downs, Def_Total_Yards, Def_Passing_Yards, Def_Rushing_Yards, Turnovers_Gained)
head(def_vars)
```

```{r}
nfl_cleaned$Win_Loss <- as.factor(nfl_cleaned$Win_Loss)

cor(nfl_cleaned$Off_Total_Yards,
    nfl_cleaned$Off_Passing_Yards + nfl_cleaned$Off_Rushing_Yards)
```

We remove Total_Yards because this variable is perfectly collinear with Passing_Yards and Rushing_Yards. This is because Total_Yards = Passing_Yards + Rushing_Yards.

```{r}
off_log <- glm(Win_Loss ~ Points_Scored + Off_First_Downs + Off_Passing_Yards + Off_Rushing_Yards + Turnovers_Lost,
               data = nfl_cleaned, family = binomial)

off_log
summary(off_log)
```

```{r}
def_log <- glm(Win_Loss ~ Points_Allowed + Def_First_Downs + Def_Passing_Yards + Def_Rushing_Yards + Turnovers_Gained,
               data = nfl_cleaned, family = binomial)

def_log
summary(def_log)
```

We are going to remove Points_Scored and Points_Allowed to remove perfect separation.

```{r}
all_log <- glm(Win_Loss ~ Off_First_Downs + Off_Passing_Yards + Off_Rushing_Yards + Turnovers_Lost + 
                 Def_First_Downs + Def_Passing_Yards + Def_Rushing_Yards + Turnovers_Gained, 
               data = nfl_cleaned, family = binomial)

all_log
summary(all_log)
```

These results show us that the way the dataset is formatted, more specifically how both sides of each game are recorded, we are having difficulties predicting wins and losses efficiently. Therefore, we are going to take different approaches.

```{r}
differences <- nfl_cleaned %>%
  select(Win_Loss) %>%
  mutate(Passing_Diff = nfl_cleaned$Off_Passing_Yards - nfl_cleaned$Def_Passing_Yards) %>%
  mutate(Rushing_Diff = nfl_cleaned$Off_Rushing_Yards - nfl_cleaned$Def_Rushing_Yards) %>%
  mutate(Turnover_Diff = nfl_cleaned$Turnovers_Gained - nfl_cleaned$Turnovers_Lost)

diff_log <- glm(Win_Loss ~ Passing_Diff + Rushing_Diff + Turnover_Diff,
                data = differences, family = binomial)

diff_log
summary(diff_log)
```

```{r}
exp(0.00929) # Passing Differences
exp(0.00929 * 50) # 50 Yard Passing Difference
exp(0.01431) # Rushing Differences
exp(0.01431 * 50) # 50 Yard Rushing Difference
exp(0.8818) # Turnover Differences

1.591 / (1 + 1.591) # turning a true 50/50 game to a +1 advantage
2.045 / (1 + 2.045)
2.415 / (1 + 2.415)
```

In an otherwise evenly matched game, having a 50 yard passing advantage over your opponent multiples the odds of winning by 1.59. This means that a 50-yard passing advantage over your opponent takes your win probability from 50% to 61.4%. 

In an otherwise evenly matched game, having a 50 yard rushing advantage over your opponent multiples the odds of winning by 2.04. This means that a 50-yard rushing advantage over your opponent takes your win probability from 50% to 67.2%.

In an otherwise evenly matched game, each additional turnover advantage multiples the odds of winning by 2.42. This means that gaining one turnover over your opponent takes your win probability from 50% to 70.7%.

```{r}
pred_prob <- predict(diff_log, type = "response")
pred_class <- ifelse(pred_prob > 0.5, 1, 0)
mean(pred_class == (differences$Win_Loss == "W"))
```

```{r}
vif(diff_log)
```

